A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-10-28 jacobian_calc_CMAME.tex
More changes to qoi section

\subsubsection{Quantities of interested} 
\label{JGAM} 
As stated in the Introduction a goal of the study was to rank CS against FD and CD on the basis of accuracy. AD was ommited from the comparison because it served as the standard of accuracy for the other methods in the absence of appropriate analaytical forms for the Jacobian associated with the system solved in the study. The assumption that AD is accurate enough to serve as a standard is supported by ADs implementation as a computerized chain rule as explained in Section~\ref{ADsubsection}.

Because it makes no sense to compare tangent-stiffness matrices from different problems, load steps that started with difference guesses, or from different iterations, it was necessary to solve one
load step and conclude each iteration within that load step by updating the displacement guess only
from the results of the AD method and to subsequently feed all four methods the same updated guess
the following iteration. This decision precluded a comparison of convergence rate, since if the
methods were allowed to solve a problem at their individual pace, differences in accuracy would
produce differences in guess updates and therefor the number and nature of iterations perfomed.
Different guesses from iterations started with different previous guesses could not be data for a
valid comparison of tangent-stiffness accuracy between methods.  However a separate convergence rate
comparison study could be done given some modifications to the materials for the study described in
this paper, and of course, the resources to perform that study.   Additionally, having each of the
methods physically operate in the same process, single or one of associated parallel, allowed the
comparison of tangent-stiffness matrices as loaded from random access memory rather than from the
hard-disk.  Loading from RAM gives the benefit of vastly greater speed and the simpler programming
compared to some other solution speculatively involving dynamic file management on files which for
one of the components of the study would be on the order of $1E10$ bytes. However, the price of
running the methods together in the same process and avoiding the use of the hard-disk was that at
least two tangent-stiffness matrices had to be stored in RAM during simulation, which meant that
special high-memory compute nodes were needed for the $1E6$ peridynamic element problem series.

The other goal was to rank the four methods, that is including AD, on the basis of speed. Instead of
convergence rate being measured, speed of iteration was measured because it could be so done at the
same time as accuracy was being measured given a single tangent-stiffness calculation.  It was
assumed that the order that each method was evaluated was unimportant, that evaluating each method
sucessively within each solver iteration did not affect their performance individually and that
speed of computation did not change with time. These assumptions allowed the test program to run the
same problem with each of the methods at effectively the same time and generate an equal volume of
data from each method. It was also assumed that the Jacobian matrix calculation time for each of the
methods did not vary based on what values the independent variables held, as they do from iteration
to iteration as the guessed equilibrium solution is updated.  This assumption allowed calculation
time measurements to be averaged over all iterations within a solution attempt. The purpose of
averaging calculation time measurements was to informally address the extraneous variables of
evaluation order and computer system load due to other user's processes or processes not associated
with the study. The question is begged: if it is assumed that the value of independent variables
had no effect on iteration speed, then why couldn't each of the method have been run independently
so that convergence data were recorded, so long as system being solved was the same? The answer is
that this would have made accuracy comparisons invalid if measured concurrently with convergence
rate, while it would cost additional time and resources if convergence rate were measured in
serparate runs. Instead, while not part of the study, the example programs available on the
corresponding author's website allow the reader to make a comparison of the methods for themselves
on the basis of accuracy and convergence rate for two example nonlinear systems. In the examples,
notice how convergence rate decreases when the step size selected for forward-difference or
complex-step is set to larger numbers such that accuracy relative to AD is decreased.

>>>>

\subsubsection{Quantities of interest} 
\label{JGAM} 
As stated in the Introduction, a goal of the study was to compare CS, FD, and CD on the basis of accuracy. AD was ommited from the comparison because it served as the standard of accuracy for the other methods in the absence of appropriate analaytical forms for the tangent-stiffness matrix associated with the system solved in the study. The assumption that AD is accurate enough to serve as a standard is supported by ADs implementation as a computerized chain rule as explained in Section~\ref{ADsubsection}.

It would be a poor comparative study to compare tangent-stiffness matrices from different problems, load steps that start with different current configurations, or from different iterations; because of this, it was necessary to solve one
load step and conclude each Newton iteration within that load step by updating the displacement iterate with only the results of the AD method and to subsequently feed all four methods the same updated displacement in the following iteration. This decision precluded a comparison of Newton iteration convergence rate, since if the methods were allowed to solve a problem at their individual pace, differences in accuracy would produce differences in guess updates and therefore the number and nature of Newton iterations performed (e.g. lower accuracy predictions of the algorithmically correct tangent-stiffness could cause a loss of quadratic convergence).  Different guesses from iterations started with different previous guesses could not be data for a valid comparison of tangent-stiffness accuracy between methods.  Additionally, having each of the methods physically operate in the same process, serial or one of associated parallel, allowed the comparison of tangent-stiffness matric calculation time as cached from random access memory rather than from the hard-disk.  Caching from RAM gives the benefit of vastly greater speed and the simpler programming compared to some other solution speculatively involving dynamic file management on files which for one of the components of the study would be on the order of 10 terabytes in size. However, the price of running the methods together in the same process and avoiding the use of the hard-disk was that at least two tangent-stiffness matrices had to be stored in RAM during the simulation, which meant that special high-memory compute nodes were needed for the 1 million peridynamic node test series.

The other goal was to compare the four methods, that is including AD, on the basis of speed. Instead of convergence rate being measured, speed of iteration was measured because it could be so done at the same time as accuracy was being measured given a single tangent-stiffness calculation.  It was assumed that the order that each method was evaluated was unimportant, that evaluating each method sucessively within each solver iteration did not affect their performance individually and that speed of computation did not change with time. These assumptions allowed the test program to run the
same problem with each of the methods at effectively the same time and generate an equal volume of data from each method. It was also assumed that the tangent-stiffness matrix calculation time for each of the methods did not vary based on what values the independent variables held, as they do from iteration to iteration as the Newton solution iterate is updated.  This assumption allowed calculation time measurements to be averaged over all iterations within a solution attempt. The purpose of averaging calculation time measurements was to informally address the extraneous variables of evaluation order and computer system load due to system processes not associated with the study. The question is begged: if it is assumed that the value of independent variables had no effect on iteration speed, then why couldn't each of the method have been run independently so that convergence data were recorded, so long as system being solved was the same? The answer is that this would have made accuracy comparisons invalid if measured concurrently with convergence
rate, while it would cost additional time and resources if convergence rate were measured in serparate runs. Instead, while not part of the study, the example programs available on the
corresponding author's website allow the reader to make a comparison of the methods for themselves
on the basis of accuracy and convergence rate for two example nonlinear systems. In the examples,
notice how convergence rate decreases when the step size selected for forward-difference or
complex-step is set to larger numbers such that accuracy relative to AD is decreased.
